---
title: "About the Gun Violence Archive — Youth Gun Violence Data"
output:
  html_document:
    theme: null
    highlight: null
    mathjax: null
    self_contained: false
    includes:
      in_header: _head_about.html
      after_body: _scripts_about.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source("../scripts/site_layout.R")

v <- format(Sys.time(), "%Y%m%d%H%M%OS3", tz = "UTC")
v <- gsub("[^0-9]", "", v)

writeLines(sprintf(
  '<meta name="description" content="Where this site\'s data comes from, how the Gun Violence Archive defines victims and perpetrators, and what the data can and cannot tell you." />\n<link rel="stylesheet" href="css/style.css?v=%s" />',
  v
), "_head_about.html")

# No charts on this page — no chart JS needed.
writeLines("", "_scripts_about.html")
```

```{r nav, results='asis', echo=FALSE}
cat(nav_html("about"))
```

<header class="site-header page-header">
<div class="wrap">
<h1>About the Gun Violence Archive</h1>
</div>
</header>

<main class="wrap article">

The [Gun Violence Archive](https://www.gunviolencearchive.org/){target="_blank" rel="noopener"} (GVA) is an independent, nonprofit research group that has tracked gun violence incidents across the United States in near-real time since 2013. GVA compiles incident records from thousands of sources — police reports, government agencies, and media accounts — and verifies each one before adding it to its database. It's one of the most complete public records of American gun violence available, and it's the primary data source behind everything on this site.

The data behind this site was shared with us by [The Trace](https://projects.thetrace.org/gun-violence-map/){target="_blank" rel="noopener"}, a nonprofit newsroom that covers gun violence in America. You can explore the underlying GVA data in more detail on their interactive map.

## How we define victims and perpetrators

GVA logs each person touched by a gun incident individually — as a victim, a perpetrator, or both — along with their age or age-group. We used those individual records to build the charts and dashboard on this site, aggregating up to national and state-level rates rather than reporting on individual incidents.

"Victim" is defined broadly, following GVA's own methodology: it includes anyone shot, anyone injured in a firearm assault, anyone caught up in an armed incident that didn't result in injury (robbery, burglary, home invasion, road rage), and victims of suicide. Most of the victimizations referenced on this site involved someone being shot, but the broader definition means the victim counts here capture more than shootings alone.

"Perpetrator" means someone GVA's records identify as responsible for a gun incident — including, in some cases, someone arrested only for illegally possessing a firearm, with no other offense attached. Because possession-only arrests can behave very differently from offenses involving an actual shooting or threat, every perpetrator rate on this site excludes them unless we say otherwise, and we label possession-only figures separately wherever we show them.

## Age brackets

GVA categorizes people under 18 in two groups: children (ages 0–11) and youth (ages 12–17). Some charts on this site show children and youth combined, in which case we say so explicitly.

## Rates and population estimates

Wherever this site reports a rate — per 100,000 youth, per 100,000 children, or per 100,000 residents — the population denominator comes from [CDC SEER population estimates](https://seer.cancer.gov/popdata/download.html), broken out by state, year, and age. National rates sum the underlying state-level estimates.

Where a chart shows an "annual rate," we've either summed 12 months of incidents directly or annualized a single month or a trailing 12-month average by multiplying it by 12. Either way, the goal is a rate expressed per year rather than per month, since a single month's rate on its own is a very small, hard-to-interpret number.

## What this data can't tell you

GVA's data is a record of gun violence incidents that were publicly known. These include incidents that were publicized by the police, a government agency, or the media; incidents that never generate a public record aren't captured. That means changes in reporting practices, policing, or media coverage can move these numbers independent of any real change in gun violence itself. We think GVA is among the best sources available for tracking gun violence over time, but like any dataset built from public records, it reflects what gets reported as much as what happens.

Ready to see what the data actually shows? Start with the [key takeaways](takeaways.html), or go straight to the [dashboard](dashboard.html) to explore it yourself.

</main>

<footer class="site-footer">
<div class="wrap">
<p>Data: [Gun Violence Archive](https://www.gunviolencearchive.org/){target="_blank" rel="noopener"}. Population estimates: CDC SEER.</p>
</div>
</footer>
