---
title: "State-Level Dashboard — Youth Gun Violence"
output:
  html_document:
    theme: null
    highlight: null
    mathjax: null
    self_contained: false
    includes:
      in_header: _head_dashboard.html
      after_body: _scripts_dashboard.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source("../scripts/site_layout.R")

v <- format(Sys.time(), "%Y%m%d%H%M%OS3", tz = "UTC")
v <- gsub("[^0-9]", "", v)

writeLines(sprintf(
  '<meta name="description" content="Explore Gun Violence Archive youth and child gun violence data by state, age group, and rate." />\n<link rel="stylesheet" href="css/style.css?v=%s" />',
  v
), "_head_dashboard.html")

# dashboard.js is the only script that calls d3.* directly (via sequentialRamp
# in utils.js and the choropleth code) and needs topojson to draw the map.
writeLines(sprintf('<script src="https://cdn.jsdelivr.net/npm/d3@7.9.0/dist/d3.min.js"></script>
<script src="https://cdn.jsdelivr.net/npm/@observablehq/plot@0.6.16/dist/plot.umd.min.js"></script>
<script src="https://cdn.jsdelivr.net/npm/topojson-client@3.1.0/dist/topojson-client.min.js"></script>
<script src="js/utils.js?v=%1$s"></script>
<script src="js/dashboard.js?v=%1$s"></script>', v), "_scripts_dashboard.html")
```

```{r nav, results='asis', echo=FALSE}
cat(nav_html("dashboard"))
```

<header class="site-header page-header">
<div class="wrap">
<h1>State-Level Dashboard</h1>
</div>
</header>

<section class="dashboard-section">
<div class="wrap">
<p class="dashboard-note">Explore trends in any state. This dashboard allows you to switch between counts and rates, choose whether to view raw monthly data or a 12-month rolling average, and explore trends for youth (age 12-17), children (age 0-11), or all children under age 18. This dashboard also allows you to include or exclude incidents that involved only a person under age 18 posessing a firearm. All figures come from Gun Violence Archive. Rates are an annual rate per 100,000 residents in the relevant age group, using CDC SEER single-age population estimates. Rolling averages, where selected, are calculated as a trailing 12-month mean.</p>

<h2>Map of Child and Youth Involvement in Gun Violence, March 2025 through February 2026</h2>
<div class="dashboard-controls">
<div class="control">
<label for="ctl-map-metric">Map metric</label>
<select id="ctl-map-metric">
<option value="youth_victim">Youth (12–17) victimization</option>
<option value="youth_perp_minus_possession">Youth (12–17) offenses (excl. possession)</option>
<option value="child_victim">Children (0–11) victimization</option>
<option value="child_perp_minus_possession">Children (0–11) offenses (excl. possession)</option>
</select>
</div>
<div class="control">
<label for="ctl-metric-type">View</label>
<select id="ctl-metric-type">
<option value="rate">Rate per 100,000</option>
<option value="count">Raw counts</option>
</select>
</div>
</div>

<figure class="chart">
<div id="dash-map"></div>
<figcaption id="dash-map-caption"></figcaption>
</figure>
<br>

<div class="dashboard-controls">
<div class="control">
<label for="ctl-state">State</label>
<select id="ctl-state"></select>
</div>
<div class="control">
<label for="ctl-population">Age range</label>
<select id="ctl-population">
<option value="youth">Youth (12–17)</option>
<option value="child">Children (0–11)</option>
<option value="">Children and youth (0–17)</option>
</select>
</div>
<div class="control">
<label for="ctl-status">Show</label>
<select id="ctl-status">
<option value="both">Victims and perpetrators</option>
<option value="victim">Victims only</option>
<option value="perp">Perpetrators only</option>
</select>
</div>
<div class="control">
<label for="ctl-possession">Offense filter</label>
<select id="ctl-possession">
<option value="all">All offenses</option>
<option value="excl">Exclude possession-only</option>
<option value="only">Possession-only only</option>
</select>
</div>
<div class="control">
<label for="ctl-metric-type-trend">View</label>
<select id="ctl-metric-type-trend">
<option value="rate">Rate per 100,000</option>
<option value="count">Raw counts</option>
</select>
</div>
<div class="control toggle-control">
<label class="toggle">
<input type="checkbox" id="ctl-smoothing" checked />
<span>12-month rolling average</span>
</label>
</div>
</div>

<div class="dashboard-controls">
<div class="control comparison-control">
<label for="ctl-compare">Compare up to 3 more states</label>
<select id="ctl-compare" multiple size="4"></select>
<p class="hint">Cmd/Ctrl-click to select multiple. "United States" is listed first in both dropdowns if you want the national trend.</p>
</div>
<div class="control clear-control">
<button type="button" id="ctl-compare-clear">Clear comparisons</button>
</div>
</div>

<figure class="chart">
<div id="dash-trend"></div>
<figcaption>Selected state(s) and any comparisons. "United States" behaves like any other selection — pick it explicitly to see the national trend.</figcaption>
</figure>

<p>Want the headline findings first? Read the <a href="takeaways.html">key takeaways</a>. Curious about the data itself? See <a href="about.html">about the Gun Violence Archive</a>.</p>
</div>
</section>

<footer class="site-footer">
<div class="wrap">
<p>This website was published by Empact Solutions in collaboration with and funding from The Annie E. Casey Foundation. Thank you to The Gun Violence Archive for collecting this data and [The Trace](https://projects.thetrace.org/gun-violence-map/) for making it available to us.</p>
</div>
</footer>
