---
title: "Put Aside the Narratives About Young People and Gun Violence. Let’s Look At the Facts."
output:
  html_document:
    theme: null
    highlight: null
    mathjax: null
    self_contained: false
    includes:
      in_header: _head_index.html
      after_body: _scripts_index.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source("../scripts/site_layout.R")

# Cache-bust our own CSS on every render, so edits always show up on reload —
# without this, browsers (and any caching proxy in between) can keep serving
# a stale copy of these files indefinitely at their bare URL.
v <- format(Sys.time(), "%Y%m%d%H%M%OS3", tz = "UTC")
v <- gsub("[^0-9]", "", v)

writeLines(sprintf(
  '<meta name="description" content="Trends in youth gun violence" />\n<link rel="stylesheet" href="css/style.css?v=%s" />\n%s',
  v, analytics_html()
), "_head_index.html")

# No chart JS needed on the home page.
writeLines("", "_scripts_index.html")
```

```{r nav, results='asis', echo=FALSE}
cat(nav_html("home"))
```

<header class="site-header">
<div class="wrap">
<p class="kicker">Data brief</p>
<h1>Put Aside the Narratives About Young People and Gun Violence. Let’s Look At the Facts.</h1>
<p class="byline"></p>
</div>
</header>

<main class="wrap article">

Americans of all stripes want a bright future for our children, and we all want them to navigate the transition to adulthood safely. Because we care about our children, we worry about them. One of the things we worry about the most is crime: we want to protect children both from becoming victims of crime, and from perpetrating crimes against others. We worry especially about violent crime, because we know that - in a society with the world’s highest rate of civilian gun ownership - involvement in violence can be life-shattering, and even deadly. 

Addressing these issues effectively means using research and evidence to inform our public policy choices, doing more of what works (and less of what doesn’t work) to keep young people and communities safe and able to thrive. It means listening to and empowering the young people and communities who suffer the most harm from crime, violence, and guns. And most critically, it means resisting the temptation to rely on stereotypes, preconceptions, and attention-grabbing anecdotes - rather than on timely, relevant, and reliable data - to shape our understanding and guide our policy choices. Taking young people and gun violence seriously has to begin with looking clearly at the facts.

Unfortunately, that has been harder than it sounds - and harder than it should be - for a very long time. Americans’ perceptions about youth, guns, and violence have been driven more by our fears than by the facts, more by anecdotes than by data. We mustn’t settle for that, especially those of us who guide public policy and shape public opinion. We all have a responsibility to do better.

That responsibility falls in a particular way on those of us who work in the youth justice field. We can’t afford to passively tolerate the persistence of misguided, data-free narratives about the young people, families, and communities we serve. And we no longer have to. New sources of data enable us to build data-driven understandings of youth violence. We can see local, regional, and national trends in violence, and identify whether youth are primarily impacted as perpetrators or as victims.

The remainder of this site examines data about youth violence shared with us from the Gun Violence Archive by The Trace. You can explore the data in greater detail at [The Trace](https://projects.thetrace.org/gun-violence-map/). As you read and explore this data, we encourage you to think about how you might use it to inform or guide your work.

<section class="link-cards">
<a class="card" href="takeaways.html">
<h3>Key Takeaways</h3>
<p>What the Gun Violence Archive data actually shows: youth are affected far more as victims than as perpetrators, and youth are a small share of all gun violence overall.</p>
<span class="card-cta">See the data →</span>
</a>
<a class="card" href="dashboard.html">
<h3>Explore the Dashboard</h3>
<p>Dig into any state, any age group, counts or rates — build your own view of the trends.</p>
<span class="card-cta">Open the dashboard →</span>
</a>
<a class="card" href="about.html">
<h3>About the Gun Violence Archive</h3>
<p>Where this data comes from, how it’s collected, and how we defined victims, perpetrators, and rates.</p>
<span class="card-cta">Read the methodology →</span>
</a>
</section>


</main>

<footer class="site-footer">
<div class="wrap">
<p>The analyses presented here are the work of [Tom Woods](https://www.aecf.org/people/tom-woods), [Cyrus O'Brien](https://www.cyrusobrien.com/), and [Jason Melchi](https://empactsolutions.org/our-team/). This website was published by Empact Solutions in collaboration with and funding from The Annie E. Casey Foundation. Thank you to The Gun Violence Archive for collecting this data and [The Trace](https://projects.thetrace.org/gun-violence-map/) for making it available to us.</p>
</div>
</footer>
