---
title: "Key Takeaways from the Data — Youth Gun Violence"
output:
  html_document:
    theme: null
    highlight: null
    mathjax: null
    self_contained: false
    includes:
      in_header: _head_takeaways.html
      after_body: _scripts_takeaways.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source("../scripts/site_layout.R")

v <- format(Sys.time(), "%Y%m%d%H%M%OS3", tz = "UTC")
v <- gsub("[^0-9]", "", v)

writeLines(sprintf(
  '<meta name="description" content="Key takeaways from Gun Violence Archive data: youth are affected far more as victims than as perpetrators, and youth gun violence is a small share of the overall picture." />\n<link rel="stylesheet" href="css/style.css?v=%s" />',
  v
), "_head_takeaways.html")

# Plot's UMD build depends on D3 being loaded globally (it is NOT
# self-contained, despite appearances) — d3 must load before Plot. This page
# never calls topojson-client directly, so that tag (dashboard-only) is omitted.
writeLines(sprintf('<script src="https://cdn.jsdelivr.net/npm/d3@7.9.0/dist/d3.min.js"></script>
<script src="https://cdn.jsdelivr.net/npm/@observablehq/plot@0.6.16/dist/plot.umd.min.js"></script>
<script src="js/utils.js?v=%1$s"></script>
<script src="js/blog-charts.js?v=%1$s"></script>
<script src="js/incidents-charts.js?v=%1$s"></script>', v), "_scripts_takeaways.html")
```

```{r nav, results='asis', echo=FALSE}
cat(nav_html("takeaways"))
```

<header class="site-header page-header">
<div class="wrap">
<h1>Key Takeaways from the Data</h1>
<p class="kicker">August 2026</p>
</div>
</header>

<main class="wrap article">

## 1. Youth and children are more involved in gun violence as victims than as perpetrators.

People under 18 are involved in gun violence far more often as victims than as perpetrators. Since 2017, 1.7 times as many youth and children have been **victimized** by gun violence as involved as perpetrators. This figure excludes victimless incidents that involve only possession of a firearm. At the height of the recent spike in gun violence, children and youth were **twice** as likely to be victims of gun violence than perpetrators. The experience of children and youth as victims of gun violence is one of the most overlooked and enduring trends in the GVA data.

#### Ratio of youth gun violence victims to perpetrators, by year, 2017 - 2025
<figure class="chart">
<div id="chart-victim-perp-ratio"></div>
<figcaption>Youth (ages 12–17) gun violence victims per perpetrator, by calendar year — raw counts, not rates. Perpetrator counts exclude possession-only offenses. The dashed line marks the average ratio across all nine full years shown; 2026 is excluded as a partial year (data through February only). Source: Gun Violence Archive.</figcaption>
</figure>

## 2. The pandemic-era spike in gun violence was real, and youth were primarily impacted as victims. 

Around the time of the pandemic, the youth gun victimization rate doubled, from about than 10 per 100k per year in February 2019 to more than 20 in June 2023 (measured by the rolling 12-month average). The rate at which youth were involved in gun violence as perpetrators also increased significantly during this time frame, but the magnitude was far smaller. Where youth victimization per 100,000 increased by 10 youth per year, perpetration rates increases by just 3 youth per 100,000 per year - nearly a 50% increase but a much smaller magnitude. According to the GVA data, the increase in youth **victimization** by firearms is the biggest story to tell about changing gun violence around the time of the pandemic. Notably, in the media and public imagination, this victimization story was often misconstrued as a story about youth violence being out of control.

<div class="chart-controls">
<label class="toggle">
<input type="checkbox" id="smoothing-toggle" checked />
<span>Show 12-month rolling average</span>
</label>
</div>

#### Youth gun violence victims and perpetrators per year per 100,000 youth, by month, 2017 - 2026
<figure class="chart">
<div id="chart-victims-perps"></div>
<figcaption>Annualized monthly rate of youth (ages 12–17) involved in gun violence per 100,000 youth. The perpetrator rate excludes possession-only offenses.  When viewing the 12-month trailing average, each data point reflects the annual rate for the 12 months ending at that date. Source: Gun Violence Archive; population denominators from CDC SEER.</figcaption>
</figure>


## 3. The pandemic-era surge is over.

The pandemic-era surge in victimization and perpetration is over. Youth and child involvement in gun violence has fallen substantially from its peak — for victims and perpetrators alike. Children's involvement in gun violence is down approximately 40%. Youth involvement in gun violence as victims is down by nearly one-third, and youth involvement in perpetrating gun violence is down 19%. 

<div class="stat-cards-intro">
<p>Change from each group's pandemic-era peak (12-month average, 2020&ndash;2024) to the most recent 12-month average.</p>
</div>

<div class="stat-cards">
<div class="stat-card is-victim">
<p class="eyebrow">Children, ages 0&ndash;11 &middot; Victims</p>
<p class="figure">-41%</p>
<p class="statement">change in child gun violence victims</p>
<p class="detail">Children injured or killed by gun violence have fallen 41% from the pandemic-era peak to the most recent 12-month average.</p>
</div>
<div class="stat-card is-perp">
<p class="eyebrow">Children, ages 0&ndash;11 &middot; Perpetrators</p>
<p class="figure">-40%</p>
<p class="statement">change in child perpetrators of gun violence</p>
<p class="detail">The number of children committing gun violence has fallen 40% from the pandemic-era peak, excluding possession-only offenses.</p>
</div>
<div class="stat-card is-victim">
<p class="eyebrow">Youth, ages 12&ndash;17 &middot; Victims</p>
<p class="figure">-34%</p>
<p class="statement">change in youth gun violence victims</p>
<p class="detail">Youth injured or killed by gun violence have fallen 35% from the pandemic-era peak to the most recent 12-month average.</p>
</div>
<div class="stat-card is-perp">
<p class="eyebrow">Youth, ages 12&ndash;17 &middot; Perpetrators</p>
<p class="figure">-19%</p>
<p class="statement">change in youth perpetrators of gun violence</p>
<p class="detail">The number of youth committing gun violence has fallen 19% from the pandemic-era peak, excluding possession-only offenses.</p>
</div>
</div>

<p class="stat-cards-source">Source: percent change from each series' pandemic-era peak (12-month average, largest value between 2020 and 2024) to the most recent month. Perpetrator figures exclude possession-only offenses.</p>


## 4. Youth violence is a small part of the picture compared to adults.

Of all the gun violence incidents recorded in the United States, the vast majority does not involve people under 18 at all. **Across the full 2017–2026 period, about 85% of incidents involve only adults.** Gun violence is overwhelmingly an adult phenomenon. This chart shows the total number of incidents involving adults alone versus people under age 18 as victims or perpetrators. The number of adult-alone incidents increased 60% over its pre-pandemic level, where children- or youth-involved incidents increased a relatively modest 34%.


<figure class="chart">
<div id="chart-incident-volume"></div>
<figcaption>All incidents nationally, split by whether a child or youth (ages 0–17) was involved as a victim, as a perpetrator (excluding possession-only offenses), or not at all. 6-month rolling sum. Sources: Gun Violence Archive national incident totals; age-coded incident records.</figcaption>
</figure>

Gun violence in the United States is a significant public safety problem and one that became especially acute in the early part of this decade. But before, during, and after the recent increase, the shape of gun violence in the United States has overwhelmingly been defined by incidents that involved adults alone.

Want to see this broken down by state? Head to the [dashboard](dashboard.html). Curious where this data comes from? Read [about the Gun Violence Archive](about.html).

</main>

<footer class="site-footer">
<div class="wrap">
<p>Data: [Gun Violence Archive](https://www.gunviolencearchive.org/){target="_blank" rel="noopener"}. Population estimates: CDC SEER.</p>
</div>
</footer>
